---
title: Example - Scottish Lip Cancer
author: "Mine Çetinkaya-Rundel & Colin Rundel"
date: "June 24, 2018"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
exclude: true

```{r setup, include=FALSE}
options(width = 80, tibble.width = 80)

knitr::opts_chunk$set(
  fig.align = "center"
)

library(magrittr)
library(ggplot2)
library(Rcpp)
library(sf)
```

---

## The data


```{r print-data}
lip_cancer = readRDS("results/lip_cancer.rds")
lip_cancer
```

---
class: middle

.pull-left[
```{r plot-obs, echo=FALSE, warning=FALSE}
lip_cancer %>% dplyr::transmute(log(Observed)) %>% plot(asp=1, breaks=seq(0,5,0.5), axes=TRUE)
```
]

.pull-right[
```{r plot-exp, echo=FALSE}
lip_cancer %>% dplyr::transmute(log(Expected)) %>% plot(asp=1, breaks=seq(0,5,0.5), axes=TRUE)
```
]

---
class: middle

```{r plot-obs-exp, echo=FALSE}
lip_cancer %>% dplyr::transmute(Observed/Expected) %>% plot(asp=1, axes=TRUE)
```

---

```{r neighbors, echo=FALSE, warning=FALSE}
m = readRDS("results/m.rds")
W = readRDS("results/W.rds")

lip_cancer$n_neighbors = m

plot(lip_cancer[,"n_neighbors"], asp=1, axes=TRUE)
```

---

## A hierachical model for lip cancer 

We have observed counts of lip cancer for 56 districts in Scotland. Let $y_i$ represent the number of lip cancer for district $i$.

\begin{aligned}
y_i &\sim \text{Poisson}(\lambda_i) \\
\\
\log(\lambda_i) &= \log(E_i) + x_i \beta + \omega_i \\
\\
\boldsymbol{\omega} &\sim \mathcal{N}(\boldsymbol{0},~[\tau(\boldsymbol{D}-\phi\,\boldsymbol{W})]^{-1})
\end{aligned}


where $E_i$ is the expected counts for each region (and serves as an offet) and the parameters of the CAR are defined as follows:

* $\tau$ is the spatial precision,
* $D = \mathit{diag}(\boldsymbol{m})$, where $\{m\}_i = $ number of neighbors of location i,
* $\phi$ is the spatial "smoothness" parameter, bounded between [0,1),
* $W$ is the adjacency matrix for all locations.


.footnote[
*We have based the above and following off of Max Joseph's Exact sparse CAR models in Stan which is available [here](http://mc-stan.org/users/documentation/case-studies/mbjoseph-CARStan.html).*
]

---

## A stan model

```{r stan-model}
cat(readLines("models/car_model.stan",warn = FALSE), sep="\n")
```

---

```{r stan-params, echo=FALSE}
full_fit = readRDS("results/car_fit.rds")
print(full_fit, pars = c('beta', 'tau', 'phi', 'lp__'))
```

---

```{r stan-traceplots, fig.width=12}
rstan::traceplot(full_fit, pars = c('beta', 'tau', 'phi', 'lp__'))
```

