---
title: Example - Scottish Lip Cancer
author: "Mine Çetinkaya-Rundel & Colin Rundel"
date: "June 24, 2018"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
exclude: true

```{r setup, include=FALSE}
options(width = 80, tibble.width = 80)

knitr::opts_chunk$set(
  fig.align = "center"
)

library(magrittr)
library(ggplot2)
library(Rcpp)
```

---
class: middle
count: false

# Scottish Lip Cancer Example

---

## The data

```{r get-data, include=FALSE}
## Get data
datafile = "http://web1.sph.emory.edu/users/lwaller/book/ch2/scotland.dat"

col_names = read.table(datafile, nrows=1, sep=" ", stringsAsFactors = FALSE) %>% 
  dplyr::select(2:7) %>%
  unlist()

data = read.table(datafile, header=FALSE, skip=1) %>%
  setNames(col_names)


## Get shapefile(s)

shape_dir = tempdir()

base_url = "http://web1.sph.emory.edu/users/lwaller/book/ch9/"
shapefiles = c("scot.shp", "scot.dbf", "scot.shx")
for(file in shapefiles) {
  download.file(file.path(base_url, file), destfile = file.path(shape_dir, file), quiet = TRUE)
}

## Combine

lip_cancer = sf::read_sf(shape_dir) %>%
  sf::st_set_crs("+proj=longlat +datum=NAD27 +no_defs") %>%
  dplyr::rename(Name = NAME, District = ID) %>% 
  dplyr::full_join(data) %>%
  dplyr::rename(id = District, District = Name, AFF = `%AFF`)
```

```{r print-data}
lip_cancer
```

---
class: middle

.pull-left[
```{r plot-obs, echo=FALSE, warning=FALSE}
lip_cancer %>% dplyr::transmute(log(Observed)) %>% plot(asp=1, breaks=seq(0,5,0.5), axes=TRUE)
```
]

.pull-right[
```{r plot-exp, echo=FALSE}
lip_cancer %>% dplyr::transmute(log(Expected)) %>% plot(asp=1, breaks=seq(0,5,0.5), axes=TRUE)
```
]

---
class: middle

```{r plot-obs-exp, echo=FALSE}
lip_cancer %>% dplyr::transmute(Observed/Expected) %>% plot(asp=1, axes=TRUE)
```

---

```{r neighbors, echo=FALSE, warning=FALSE}
d = sf::st_distance(lip_cancer)
class(d) = NULL

W = (d == 0.0) * 1L

m = rowSums(W)
lip_cancer$n_neighbors = m

plot(lip_cancer[,"n_neighbors"], asp=1, axes=TRUE)
```

---

## A hierachical model for lip cancer 

We have observed counts of lip cancer for 56 districts in Scotland. Let $y_i$ represent the number of lip cancer for district $i$.

\begin{aligned}
y_i &\sim \text{Poisson}(\lambda_i) \\
\\
\log(\lambda_i) &= \log(E_i) + x_i \beta + \omega_i \\
\\
\boldsymbol{\omega} &\sim \mathcal{N}(\boldsymbol{0},~[\tau(\boldsymbol{D}-\phi\,\boldsymbol{W})]^{-1})
\end{aligned}


where $E_i$ is the expected counts for each region (and serves as an offet) and the parameters of the CAR are defined as follows:

* $\tau$ is the spatial precision,
* $D = \mathit{diag}(\boldsymbol{m})$, where $\{m\}_i = $ number of neighbors of location i,
* $\phi$ is the spatial "smoothness" parameter, bounded between [0,1),
* $W$ is the adjacency matrix for all locations.


.footnote[
*We have based the above and following off of Max Joseph's Exact sparse CAR models in Stan which is available [here](http://mc-stan.org/users/documentation/case-studies/mbjoseph-CARStan.html).*
]

---

## A stan model

```{r stan-model}
car_model = "
  data {
    int<lower = 1> n;
    int<lower = 1> p;
    matrix[n, p] X;
    int<lower = 0> y[n];
    vector[n] log_offset;
    matrix<lower = 0, upper = 1>[n, n] W;
    vector[n] m;
  }
  transformed data{
    matrix<lower = 0>[n, n] D;
    D = diag_matrix(m);
  }
  parameters {
    vector[p] beta;
    vector[n] omega;
    real<lower = 0> tau;
    real<lower = 0, upper = 1> phi;
  }
  model {
    omega ~ multi_normal_prec(rep_vector(0, n), tau * (D - phi * W));
    beta ~ normal(0, 1);
    tau ~ gamma(2, 2);
    y ~ poisson_log(log_offset + X * beta + omega);
  }
"
```

---

```{r stan-fit, results="hide"}
data = list(
  n = nrow(lip_cancer), p = 1,
  W = W, m = m,
  log_offset = log(lip_cancer$Expected),
  y = lip_cancer$Observed,
  X = matrix(1, nrow = nrow(lip_cancer), ncol=1)
) 

options(mc.cores = parallel::detectCores())
full_fit <- rstan::stan(model_code = car_model, data = data, 
                        iter = 1000, chains = 2, verbose = FALSE)
```

---

```{r stan-params}
print(full_fit, pars = c('beta', 'tau', 'phi', 'lp__'))
```

---

```{r stan-traceplots, fig.width=12}
rstan::traceplot(full_fit, pars = c('beta', 'tau', 'phi', 'lp__'))
```

